---
title: "ASLO Montreal: R Workshop"
author: "Kateri Salk"
date: "2026-05-13"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Limnology in Montreal

We're excited to dig into some limnology coding here in Montreal! Whether you're a beginner R user or want to further hone your R coding skills, you're in the right place. We'll focus on a Canadian lake dataset, through which we'll demonstrate a data processing workflow and explore limnology concepts.

### Objectives

1.  Write code and documentation in R scripts
2.  Create code to process and analyze data
3.  Perform exploratory analyses and visualization on lake datasets

### Dataset

Today, we'll be using a dataset from the IISD-Experimental Lakes Area (IISD-ELA). This field station, located in northwest Ontario, is home to some of the most impactful lake research in the world. IISD-ELA has operated a long-term monitoring program for nearly 60 years and has also conducted whole-ecosystem experiments to inform knowledge and policy on lake stressors, including nutrients, acidification, pharmaceuticals, heavy metals, food web alteration, reservoir development, and climate change.

For more information, see these links:

<https://www.iisd.org/ela/visit-the-lakes/our-lakes/>

<https://github.com/IISD-ELA/info-sheets/tree/main>

## Introduction to our Coding Setup

### RStudio Basics

Welcome to the RStudio interface. When you open RStudio, you will see four panels:

**Source Code Editor** (top left) includes a tab structure to pull up and edit R scripts and markdown documents.

**Console** (bottom left) interacts with R processes. R code is run here. There is also a tab here called `Terminal` which will allow you to access git functionality.

**Workspace Browser** (top right) holds the `global environment` that is populated by analyses run in each R session. There is also a `history` tab and a `git` tab.

**Notebook** (bottom right) holds tabs for `files`, `plots`, `packages`, and `help`. You will interact with each of these functionalities, and we will explain each as they come up.

More on the functionality of each of these panels as we move through this lesson.

### Quarto documents

You are currently viewing a Quarto document. This type of file includes text chunks and R code chunks that can be viewed together. Quarto documents can also be exported ("knitted") into a PDF or html format. Quarto documents are the successor to RMarkdown documents and are functionally the same for today's purposes. For more on that, see: [this article](https://quarto.org/docs/faq/rmarkdown.html).

An R script file is similar to a Quarto document, except it *only* includes R code. Any text that is included in an R script that it not intended to be run as R code must be "commented out" so that R does not interpret the text as code. We'll use comments as well today.

You may also choose to type or paste R code directly into the console. This is not a recommended method, as it undermines the goals of reproducibility (code is not saved). However, typing directly into the console can be useful if you need to do something that is strictly temporary (e.g., run a quick calculation, look at a summary of a dataset).

We've organized this Quarto document as an outline (see upper right portion of the console). The pound/hashtag symbols indicate the outline levels for easy referencing and navigation.

### R Coding basics

Below is a chunk of R code. You can run R code in several ways:

-   Place your cursor on the line of R code that you want to run, then press `control + enter` (PC) or `command + enter` (Mac). Your R code should appear in the console, followed by any output generated by the code.

-   Highlight line(s) of R code, then press `control + enter` (PC) or `command + enter` (Mac). Your R code should appear in the console, followed by any output generated by the code. This is a good option if you want to run multiple lines of code at once.

You may also choose to type or paste R code directly into the console. This is not a recommended method, as it undermines the goals of reproducibility (code is not saved). However, typing directly into the console can be useful if you need to do something that is strictly temporary (e.g., look at a summary of a dataset or determine the class of a variable).

You can create R objects with an *assignment* statement. The indicator for an assigment is the `<-` symbol. A good way to think about the meaning of an assignment statement is "object name (lefthand side) gets value (righthand side)."

A quick note: in many situations, a `=` sign will substitute for a `<-`. Resist this temptation! Many functions use the `=` symbol in a way that is not interchangeable with `<-`, so it's best to get in the habit of differentiating an *assignment* from an *equals* statement.

```{r}

x <- 3*4

# Notice that `x` has also just shown up in your Environment tab. 
# Now, call up the object `x`. 

x


```

R objects can be named with a combination of letters, numbers, underscore (`_`) and period (`.`). The best R object names are *informative*. Resist the temptation to call your R object something convenient, like "a", "b", and so on. Calling your R object something specific means that you can call up that object later and have an idea of what it contains, with less need for specific context.

Informative names are the first illustration of a common data management recommendation: take the time to use best management practices at the outset, and it will save you time in the long term.

Importantly, you may never call an R object "data". This word is reserved for a specific function and may not be assigned as a name. To work around this, many people call their R objects "dat", which is another example of a less-than-ideal data management practice because it is not informative.

### Functions

R functions are the major tool used in R. Functions can do virtually unlimited things within the R universe, but each function requires specific inputs that are provided under specific syntax. We will start with a simple function that is built into R, `seq`

```{r}
seq(1, 10)
ten_sequence <- seq(1, 10) # from, to
ten_sequence

seq(1, 10, 2) # from, to, by

# remove an object from the environment (helpful to keep environment clean)
rm(ten_sequence, x)
```

### Tips and Tricks

-   Case and spelling matter

-   Spaces (generally) don't matter. One notable exception is that spaces within quotation marks *do* matter.

-   Parentheses and quotation marks appear in pairs when typed into RStudio.

-   When typing long names, use the `tab` key partway through the name to generate autocomplete options.

-   In the upper right corner of the editor is a button with multiple horizontal lines. Clicking this button will bring up the outline of the document. Headings in the outline are determined by how you've defined them in the document.

## Setup

### Load packages and data

First, we will load a few packages we will use today, which include functions that will help explore, process, and visualize the dataset. Then, we will upload the dataset that contains the Experimental Lakes Area long-term data.

```{r}
#| message: false
#| warning: false

# call our first functions: install and load function packages

# install.packages("tidyverse")
# install.packages("lubridate")
# install.packages("viridis")

library(tidyverse)
library(lubridate)
library(viridis)

# Check our working directory. 
getwd()

# read in the dataset as an R object.
# tip: don't call your dataset "data"! Always choose something informative.
eladata <- read.csv("ELA_dataset.csv")

```

## Initial dataset exploration

First, we'll take a look at the dataset. These several functions use a `function(dataset)` syntax, and they are useful for getting a quick understanding of the different columns of the dataset and what they contain.

```{r}

class(eladata)
colnames(eladata)
view(eladata)
summary(eladata)
head(eladata)
glimpse(eladata)

```

## Data Processing

### Formatting dates

R will often import dates as factors or characters rather than dates. To fix, this we need to tell R that it is looking at dates. We also need to specify the format the dates are in. By default, if you don't provide a format, R will attempt to use %Y-%m-%d or %Y/%m/%d as a default. Note: if you are working collaboratively in an international setting, using a year-month-day format in spreadsheets is the least ambiguous of date formats. Make sure to check whether month-day-year or day-month-year is used in an ambiguously formatted spreadsheet.

Formatting of dates in R:

-   %d day as number (0-31)
-   %m month (00-12, can be e.g., 01 or 1)
-   %y 2-digit year
-   %Y 4-digit year
-   %b abbreviated month
-   %B unabbreviated month

In some cases when dates are provided as integers, you may need to provide an origin for your dates. Beware: the "origin" date for Excel (Windows), Excel (Mac), R, and MATLAB all have different origin dates. Google this if it comes up.

```{r}

class(eladata$activity_start_date)
eladata$activity_start_date <- as.Date(eladata$activity_start_date, format = "%Y-%m-%d")
summary(eladata$activity_start_date)

```

We may also want to add a few extra columns pertaining to dates for exploration and visualization purposes. In this case, we'll use the `tidyverse` and `lubridate` packages to add this information.

Data processing functions:

-   `mutate`: adds a column to the dataset based on a conditional statement.

-   **Pipe operator:** `%>%` or `|>`; indicates the linking of a series of functions in sequence within a single code chunk. Add this to the end of a line of code so that R will continue running the next line.

Take notes in the comments as we move along with the data processing!

```{r}

eladata <- eladata  %>% # 
  mutate(Year = year(activity_start_date), # 
         Month = month(activity_start_date), # 
         Week = week(activity_start_date)) # 


```

### Wrangling limnology data

First, we'll do some checks on what the dataset contains.

```{r}

unique(eladata$activity_media_name)
unique(eladata$activity_depth_height_unit)
unique(eladata$characteristic_name)


```

What types of data are available from the Experimental Lakes Area dataset? What types of insights might we be able to gain from these data? Are there any implications, particularly with regard to depth, that we should keep in mind as we continue processing the data?

> ANSWER:

It's good practice to keep the imported dataset largely unchanged so you don't need to go back and re-import it from the beginning if you make a mistake or change course. So, what we can do is create processed datasets that build off the main dataset but fulfill specific analysis needs.

Data processing functions:

-   `select`: includes (or excludes) certain columns from the new dataset.

-   `filter`: includes (or excludes) certain rows from the new dataset, using conditional statements.

-   `case_when`: specifies an operation to be done if the conditional statement is true. Uses a format of conditional statement, followed by a `~`, then the resulting operation.

-   `group_by`: groups the dataset by a certain set of columns so that a subsequent function can be done by the group rather than individually.

-   `summarise`: often used with `group_by`, summarizes a condition. This time, we'll calculate the mean of duplicate measurements so that the resulting dataset contains only one observation for each location, depth, date, and parameter.

-   `pivot_wider`: takes a given column and creates a set of new columns based on the factor levels in the original column. In this case, we'll take the `characteristic_name` column and create new columns, one for each parameter.

First, we'll select a subset of the columns to make the dataset easier to view. Then, we'll filter the dataset to include only a select set of parameters, for surface depths, and for the season that contains May through October.

After that, we will work to *pivot* the dataset wider, meaning that each parameter will have its own column. This can make it easier in some cases to visualize and statistically analyze the data. In order to do this, we'll need to preserve a combination of the parameter name and unit.

```{r}
#| message: false
#| warning: false

eladata_surface <- eladata %>% # 
  select(monitoring_location_name, activity_start_date,
         activity_depth_height_measure, activity_depth_height_unit,
         characteristic_name, characteristic_name_long, 
         result_value, result_unit,
         Year, Month, Week) %>% # 
  filter(characteristic_name %in% c("TDP", "PARTP", "TDN", "PARTN", 
                                    "CHLA", "DOC", "PH", "COND")) %>%
  filter(activity_depth_height_unit == "epi" | activity_depth_height_unit == "surface grab") %>%
  filter(Month %in% c(5:10)) %>%
  mutate(Parameter = case_when(characteristic_name == "TDP" ~ "TDP_ugL",
                               characteristic_name == "PARTP" ~ "PARTP_ugL",
                               characteristic_name == "TDN" ~ "TDN_ugL",
                               characteristic_name == "PARTN" ~ "PARTN_ugL",
                               characteristic_name == "CHLA" ~ "CHLA_ugL",
                               characteristic_name == "DOC" ~ "DOC_umolL",
                               characteristic_name == "PH" ~ "pH",
                               characteristic_name == "COND" ~ "COND_uScm")) %>%
  select(-characteristic_name, -characteristic_name_long, -result_unit) %>%
  group_by(monitoring_location_name, activity_start_date,
           activity_depth_height_measure, activity_depth_height_unit,
           Parameter, Year, Month, Week) %>%
  summarise(result_value = mean(result_value, na.rm = TRUE)) %>%
  pivot_wider(names_from = "Parameter", values_from = "result_value") %>%
  mutate(TP_ugL = TDP_ugL + PARTP_ugL,
         TN_ugL = TDN_ugL + PARTN_ugL)

```

Next, let's make a separate dataset focused on Lake 239, which has depth profile information. We'll use all the same functions as before, but we'll play with adding multiple conditional statements within `filter`.

```{r}
#| message: false
#| warning: false

eladata_L239 <- eladata %>%
  select(monitoring_location_name, activity_start_date,
         activity_depth_height_measure, activity_depth_height_unit,
         characteristic_name, characteristic_name_long, 
         result_value, result_unit,
         Year, Month, Week) %>%
  filter(monitoring_location_name == "239 LA CB" & 
           characteristic_name == "O2" &
           activity_depth_height_unit != "epi" & 
           Year %in% c(2015:2019)) 

```

## Data Visualization

One of the most popular data visualization packages in R is called `ggplot2`. When we load the `tidyverse` package, the `ggplot2` package is automatically loaded as a sub-package. ggplot is a flexible visualization framework that allows you to choose different graph types and also link layers of formatting into one chunk of code. Rather than using a `pipe` operator, we will instead indicate we're continuing the ggplot layers with a `+` sign.

Today, we'll be making the following common plot types:

-   Boxplot

-   Scatterplot

-   Line plot (to illustrate lake depth profiles)

-   Time series plot

### Boxplot

To start, let's use our surface dataset to look at chlorophyll *a* concentrations by month and lake. First, we specify the `ggplot` function. Within that function, we indicate the dataset that we'd like to visualize, as well as any aesthetics that we'll use across the graph (i.e., the x and y axes, colors, symbols, etc.). Then, we use a `+` sign, followed by the plot type we'd like to use. In this case, let's try a boxplot using the `geom_boxplot` function.

```{r}
#| message: false
#| warning: false

ggplot(eladata_surface,
       aes(x = as.factor(Month), y = CHLA_ugL, fill = monitoring_location_name)) +
  geom_boxplot()

```

We've successfully created a graph! However, it's not very successful in terms of being interpretable and appealing. Let's add some additional code to improve the graph. Here, we are adding in a few functions baked into the `ggplot2` package, as well as a color scale included as part of the `viridis` package.

```{r}
#| message: false
#| warning: false

ggplot(eladata_surface,
       aes(x = as.factor(Month), y = CHLA_ugL, fill = monitoring_location_name)) +
  geom_boxplot() +
  scale_y_log10() + # when the range spans orders of magnitude, this can be helpful.
  scale_fill_viridis_d(option = "plasma", begin = 0.1, end = 0.9) + #
  labs(x = "Month",
       y = expression("Chlorophyll a ("*mu*"g/L)"),
       fill = "Lake") + #
  theme_classic() + # theme_minimal is also a good one.
  theme(legend.position = "top") #


```

What insights can you draw from this graph in terms of the variability of chlorophyll across months and among lakes?

> ANSWER:

### Scatterplot

Next, let's try creating a scatterplot. We'll use the `geom_point` function to achieve this. Let's explore a common stressor-response relationship, comparing the concentrations of TP to the concentrations of chlorophyll *a*. We may want to also add a linear regression line, which we can do using the `geom_smooth` function. Note that it's always best to run a statistical test confirming the significance of the regression before adding a line of best fit to the graph (but that's a topic for another day).

```{r}
#| message: false
#| warning: false

ggplot(eladata_surface, aes(x = TP_ugL, y = CHLA_ugL, color = monitoring_location_name)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)

```

Let's improve the plot. What might you try given what we explored with the boxplot?

```{r}
#| message: false
#| warning: false

ggplot(eladata_surface, aes(x = TP_ugL, y = CHLA_ugL, color = monitoring_location_name)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) 


```

What do you notice about the relationship between TP and chlorophyll? Does this vary across lakes? (remember that this scatterplot displays correlation, not causation)

> ANSWER:

*Optional exercise:* Create a different scatterplot exploring the connection between two other variables. What might that tell you about the limnological dynamics in these lakes?

### Line graph (depth profiles)

Next, let's create a plot that illustrates a core visualization in limnology: the depth profile. In this case, we have dissolved oxygen (DO) data from Lake 239 collected at several depths from 2015 to 2019.

```{r}
#| message: false
#| warning: false

ggplot(eladata_L239, 
       aes(x = result_value, y = activity_depth_height_measure, color = as.factor(Week))) +
  geom_line(orientation = "y") + #
  geom_point() + #
  facet_wrap(vars(Year), ncol = 5) + #
  scale_y_reverse() + #
  scale_color_viridis_d(direction = -1, end = 0.9) + #
  labs(x = "DO (mg/L)", y = "Depth (m)", color = "Week") + 
  theme_classic() +
  theme(legend.position = "top")

 
```

*Optional exercise:* explore depth profiles of other parameters. What does this tell you about the biogeochemistry of Lake 239 and how it varies as the season progresses?

### Time series plot

When we have a **date** column in the dataset, we can use scatterplot or line plot functions to create a time series plot.

```{r}
#| message: false
#| warning: false

ggplot(eladata_L239, 
       aes(x = activity_start_date, y = result_value, color = activity_depth_height_measure)) +
  geom_point() +
  scale_color_viridis_c(direction = -1, end = 0.9) +
  theme_classic() +
  theme(legend.position = "top")
```

What can you interpret about how DO varies by depth and over time? Personally, do you prefer a depth profile visualization or a time series visualization?

> ANSWER:
